Physics-based protein structure prediction and design using the confinement method

### 2013-01-11 Paper/paper.tex
changed paper 11_5

Calculation of free energy have been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}. In recent years methods like the Reference system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008} or the Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} to name a few have been used to calculate free energy
differences.
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that the free energy difference is a state function and 
thus it is independent of the sampling path. This approach uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraints to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Next we have applied this method to larger proteins for the
first time. The free energy difference of pairs of conformations in proteins with similar sequence but completely
different fold was calculated. \Alberto{i made this part too long, have to summarize here and define
better inside text?-->} Encouraged by the result, we applied the method to help tackle the
problem of protein conformation ranking. We did this by selecting different targets from the
Critical Assessment of Structure Prediction (CASP) event. In this event, different research groups
try to find the 3D structure of a protein given a sequence in a blind way using their methodologies. Upon comparing
with the real structure it is often seen that several groups are able to generate better structures
than the ones they submit, but are not able to identify them. This is known as the ranking problem,
and this method is able to correctly rank structures in the majority of cases we have tried.
 Finally, we have been exploring the ability of working on a per residue basis to identify which
 residues are the major players in the stabilization/destabilization of a pair of conformations. Our
 initial experience makes us thinkg that this method might be helpful in identifying key residues
 for protein design.

@@

\subsection{Confinement method produces correct results in control experiments}

@@

\begin{table}
\begin{center}
\caption{Confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}

@@

\subsection{Confinement method correctly predicts the structural preferences of chameleon
sequences}

@@

To avoid this potential problem, we instead computed the relative free energy of two different
structural models for each sequence. One model is based on the GA structure and the other on the GB
structure (see Supporting Information for details on the modeling procedure). This is a much more
realistic test of the confinement method's ability to accurately calculate relative free energies.

@@

\subsection{Per residue free energy calculation can identify the mechanistic detail behind conformational pereference 
of a particular residue}

We decomposed the calculated free energy into per residue decomposition as described in the methods
section. This helps us to
identify important residues that stabilize a particular conformation. 
\Alberto{This is not Reader Friendly at all. we do not have to describe every little detail. I would
    rather have a table where we have the pairing and the reason for favoring one structure or the
    other. In the main text I would say that the main reason for preferences of one fold or another
    has to do with the relative solvent exposure of the residue in one conformation or another, or
    the possibility of stablishing hydrogen/salt bridges in one conformation. This would be my
preference, ditching everything below until the next section. What do you think?}
In this direction, we first calculate the per residue free energy for the GA95 and GB95 sequences. Similar to the total 
free energy for each sequence, we calculate the free energy difference between $4 \beta + \alpha$ and $3 \alpha$ structure for each 
residue. The free energy, $\Delta \Delta G (4 \beta + \alpha$ - $3 \alpha)$ is plotted in Figure~\ref{fig:orban_full}(F). 
In this plot, the negative peak indicate the residue will stabilize the $4 \beta + \alpha$ conformation of a particular
residue and the positive peak will stabilize the $3 \alpha$ structure. As, the differences of amino acid 
residues are only at 3 position, there are some common features of residues in both GA95 and GB95 sequences in both 
$3 \alpha$ and $4 \beta + \alpha$ conformation. They are reported in Figure~\ref{fig:orban_full} 
and Table~\ref{table:orban_perresidue}. Before, we further go into that detail, it is important to note that 
experimental observations classified the protein into two parts: Amino acids 9-51 are fully structured in both
the folds, where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta + \alpha$ structure.
Most of the amino acid residues in the region 1-9 favor $4 \beta + \alpha$ structure as can be seen from the plot 
Figure~\ref{fig:orban_full}(F). Here, we try to explain some of the largest peak in that plot. The residues Leu-5 and 
Leu-7 favor $4 \beta + \alpha$ structure as these hydrophobic residues are oriented towards the protein hydrophobic 
core, whereas in $3 \alpha$ structure they are exposed to the solvent (Figure~\ref{fig:orban_full}(A)). One can argue that 
hydrophilic residues in these positions can stabilize the $3 \alpha$ conformation.
The next big peak is located at position 11. A close observation of the sidechain orientation of residue Gln at position 11 reveal that it
is exposed to the protein surface in both the conformation, but in $3 \alpha$ structure it has an additional H-bond with Glu-15 which gives
it more stability compared to the $4 \beta + \alpha$ structure (Figure~\ref{fig:orban_full}(B)). The peak 
at position 14 is due to the hydrophilic Glu, which is completely
exposed to the solvent in $4 \beta + \alpha$ conformation as shown in
(Figure~\ref{fig:orban_full}(E)), but is only partially exposed 
to the solvent in $3 \alpha$ form. 
This is the reason why Glu at position 14 stabilize $4 \beta + \alpha$ conformation. Another high peak is at position 26, which indicate
that the residue in that position stabilizes $4 \beta + \alpha$ conformation. Again, the hydrophobic amino acid residue, Ala at position 26
is oriented towards the hydrophobic core of the protein in $4 \beta + \alpha$ conformation , whereas it is exposed towards the protein surface 
in $3 \alpha$ structure (Figure~\ref{fig:orban_full}(C)). Similar situations arise for residues at position 54 and 56. 
These residues are in the region, where $4 \beta + \alpha$ form is fully structured and $3 \alpha$ form is unstructured. 
As shown in Figure~\ref{fig:orban_full}(D), the sidechain of hydrophobic Val-54 is oriented towards the hydrophobic core in 
$4 \beta + \alpha$ conformation and thus have hydrophobic interactions with other residues in this conformation.
On the other hand, in $3 \alpha$ it is exposed to the solvent. The reverse happens for the residue at position 56.
Here the hydrophilic residue Glu is partially or not at all exposed to the protein surface in $4 \beta + \alpha$ but fully exposed 
to the solvent in $3 \alpha$ conformation (Figure~\ref{fig:orban_full}(G)). This is the reason, why one can see a positive peak for position 56. 
Similarly, hydrophobic Ile-49 in $4 \beta + \alpha$ conformation is exposed to protein surface and solvent, whereas as shown in 
Figure~\ref{fig:orban_full}(H)$3 \alpha$ conformation 
it has hydrophobic interaction with other hydrophobic residues in protein core.   
Now, let us discuss the differences of $3 \alpha$ and $4 \beta + \alpha$ fold due to specific difference of mutated amino acid 
residues in GA and GB sequence. The differences are in positions 20, 30 and 45. While GA95 has Leu-20-Ile-30-Leu-45, GB95 sequence 
has Ala-20-Phe-30-Tyr-45 at those position. One can find that the peak at position 49 stabilizes the GA sequence more than that of GB
sequence. In $3 \alpha$ fold, Ile-49 has hydrophobic interactions with Leu-20 and Ile-30 with GA sequence, whereas with GB sequence it 
has such interactions with smaller Ala-20 and Phe-30 residues (Figure~\ref{fig:orban_full}(I). In $3 \alpha$, with GB sequence, Phe-30 
being a larger group can not be completely accomodated in the hydrophobic core(Figure~\ref{fig:orban_full}(J)) and therefore, 
partially exposed to the protein surface. There is a very interesting difference for residue at position 45 as well. In GB95 sequence
and $4 \beta + \alpha$ fold there is H-bonding between Tyr-45 and Asp-47 (Figure~\ref{fig:orban_full}(K)), which is absent in the
$3 \alpha$ conformation. On the other hand with GA conformation, this H-bond is absent for both the conformation as there is a Leu at position
45. Although, one can not find much difference between $4 \beta + \alpha$ and $3 \alpha$ conformation
for position 20 for both GA95 and GB95 sequence. There are significant difference for the residues at position 30 and 45.  
These residues have some long term effects as well. As already explained, residue Ile-49 has much more hydrophic interactions with residues
at position 20 and 30 in GA sequence, which is less in GB sequence. There are difference for Asp-47 as well. With $3 \alpha$ conformation 
and GA95 sequence, Asp-47 has H-bonding with Lys-50, which is absent in the other conformation.

>>>>

The calculation of free energy have been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}. In recent years methods like the Reference system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008} or the Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} to name a few have been used to calculate free energy
differences.
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that the free energy difference is a state function and 
thus it is independent of the sampling path. This approach uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraints to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Next we have applied this method to larger proteins for the
first time. The free energy difference of pairs of conformations in proteins with similar sequence but completely
different fold was calculated. \Alberto{i made this part too long, have to summarize here and define
better inside text?-->} Encouraged by the result, we applied the method to help tackle the
problem of protein conformation ranking. We did this by selecting different targets from the
Critical Assessment of Structure Prediction (CASP) event. In this event, different research groups
try to find the 3D structure of a protein given a sequence in a blind way using their methodologies. Upon comparing
with the real structure it is often seen that several groups are able to generate better structures
than the ones they submit, but are not able to identify them. This is known as the ranking problem,
and this method is able to correctly rank structures in the majority of cases we have tried.
 Finally, we have been exploring the ability of working on a per residue basis to identify which
 residues are the major players in the stabilization/destabilization of a pair of conformations. Our
 initial experience makes us thinkg that this method might be helpful in identifying key residues
 for protein design.

@@

\subsection{The confinement method produces correct results in control experiments}

@@

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}

@@

\subsection{The confinement method correctly predicts the structural preferences of chameleon
sequences}

@@

To avoid this potential problem, we instead computed relative free energy of two different
structural models for each sequence. One model is based on the GA structure and the other on the GB
structure (see Supporting Information for details on the modeling procedure). This is a much more
realistic test of the confinement method's ability to accurately calculate relative free energies.

@@

\subsection{The per residue free energy calculation can identify the mechanistic detail behind conformational pereference 
of a particular residue}

We decomposed the calculated free energy into per residue decomposition as described in the method
section. This helps us to identify important residues that stabilize a particular conformation. 
In this direction, we first calculate the per residue free energy for the GA95 and GB95 sequences. 
The per residue free energy, $\Delta \Delta G (4 \beta + \alpha$ - $3 \alpha)$ is plotted in Figure~\ref{fig:orban_full}(F).
In this plot, the negative peak indicate the residue will stabilize the $4 \beta + \alpha$ conformation of a particular
residue and the positive peak will stabilize the $3 \alpha$ structure. As, the differences of amino acid
residues are only at 3 position, there are some common features of residues in both GA95 and GB95 sequences in both
$3 \alpha$ and $4 \beta + \alpha$ conformation. Role of some important residues in stabilizing either 
$4 \beta + \alpha$ or $3 \alpha$ conformation and possible reason for such conformational preferences are discussed in 
details in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}. It is important to note that
experimental observations classified the protein into two parts: Amino acids 9-51 are fully structured in both
the folds, where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta + \alpha$ structure.
Most of the amino acid residues in the region 1-9 favor $4 \beta + \alpha$ structure as can be seen from the plot
Figure~\ref{fig:orban_full}(F). Interestingly, we can explain most of the big peak in this plot. They are presented 
in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}. Here, we explain only one case. The first big peak
appears at position 7. Here, the hydrophobic Leu-7 stabilizes the $4 \beta + \alpha$ conformation as its sidechain is oriented 
towards the protein hydrophobic core, whereas in $3 \alpha$ structure it exposed to the solvent (Figure~\ref{fig:orban_full}(A)).
Thus, if someone want to better design the $3 \alpha$ conformation of this protein, we propose to mutate Leu at this position
with some hydrophilic residue. Similarly, we found that the free energy differences arise in two conformation for various 
reasons ranging from hydrogen/salt bridge formation to relative solvent exposure of hydrophobic residues etc. They are 
listed in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}.

\Alberto{This is not Reader Friendly at all. we do not have to describe every little detail. I would
    rather have a table where we have the pairing and the reason for favoring one structure or the
    other. In the main text I would say that the main reason for preferences of one fold or another
    has to do with the relative solvent exposure of the residue in one conformation or another, or
    the possibility of stablishing hydrogen/salt bridges in one conformation. This would be my
preference, ditching everything below until the next section. What do you think?}

\Arijit{I have rewrite this part.There was a table. I made it better. I want to keep some discussion though if you agree.}

Now, let us discuss the differences of $3 \alpha$ and $4 \beta + \alpha$ fold due to specific difference of mutated amino acid 
residues in GA and GB sequence. The differences are in positions 20, 30 and 45. While GA95 has Leu-20-Ile-30-Leu-45, GB95 sequence 
has Ala-20-Phe-30-Tyr-45 at those position. One can find that the peak at position 49 stabilizes the GA sequence more than that of GB
sequence. In $3 \alpha$ fold, Ile-49 has hydrophobic interactions with Leu-20 and Ile-30 with GA sequence, whereas with GB sequence it 
has such interactions with smaller Ala-20 (Figure~\ref{fig:orban_full}(I). In $3 \alpha$, with GB sequence, Phe-30 
being a larger group can not be completely accomodated in the hydrophobic core(Figure~\ref{fig:orban_full}(I)) and therefore, 
partially exposed to the protein surface. There is a very interesting difference for residue at position 45 as well. In GB95 sequence
and $4 \beta + \alpha$ fold there is H-bonding between Tyr-45 and Asp-47 (Figure~\ref{fig:orban_full}(K)), which is absent in the
$3 \alpha$ conformation. On the other hand with GA conformation, this H-bond is absent for both the conformation as there is a Leu at position
45 (Figure~\ref{fig:orban_full}(J)). Although, one can not find much difference between $4 \beta + \alpha$ and $3 \alpha$ conformation
for position 20 for both GA95 and GB95 sequence. There are significant difference for the residues at position 30 and 45.  
These residues have some long term effects as well. As already explained, residue Ile-49 has much more hydrophic interactions with residues
at position 20 and 30 in GA sequence, which is less in GB sequence. There are difference for Asp-47 as well. With $3 \alpha$ conformation 
and GA95 sequence, Asp-47 has H-bonding with Lys-50, which is absent in the other conformation.
